---
title: "Class 11 AlphaFold"
author: "Rocio Silenciario"
format: pdf
---

Here we read the results from AlphaFold and try to interpret all the models and quality score metrics:

```{r}
library(bio3d)

pth <- "dimer_23119/"
pdb.files <- list.files(path=pth, full.names = TRUE, pattern= ".pdb")
```

Align and supperpose all these models

```{r}
file.exists(pdb.files)
```

```{r}
pdbs <- pdbaln(pdb.files, fit = TRUE, exefile="msa")
```
```{r}
plot(pdbs$b[1,], typ="l", ylim=c(0,100), ylab="pLDDT score")
lines(pdbs$b[2,],typ="l", col="blue")
lines(pdbs$b[3,],typ="l", col="green")
lines(pdbs$b[4,],typ="l", col="orange")
lines(pdbs$b[5,],typ="l", col="red")
```


```{r}
pdbs$sse[1,]
```

```{r}
library(jsonlite)
pae_files <- list.files(path="dimer_23119/",
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
# Listing of all PAE JSON files
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)
attributes(pae1)
```

```{r}
#Per-residue pLDDT scores 
#same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
pae1$max_pae
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
aln_file <- list.files(path="dimer_23119/",
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```
How many sequences are in this alignment
```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the `conserv()` function.

```{r}
sim <- conserv(aln)
```

```{r}
plotb3(sim[1:99],
       ylab="Conservation Score")
```
Note the conserved Active Site residues D25, T26, G27, A28. These positions will stand out if we generate a consensus sequence with a high cutoff value:
```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```
```{r}
rd <- rmsd (pdbs)
rd

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```


